---
Title: "Homework 3"
Subtitle: "Submission 2"
Author: "Alekhya Pidugu"
format:
  pdf:
    output-file: "pidugu-a-hwk3-2"
    output-ext:  "pdf"
---
# Git Repository: https://github.com/AlekhyaPidugu/Homework3
```{r results="hide", echo=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes, fixest,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, kableExtra, modelsummary,
               kableExtra)

```
```{r  results="hide", echo=FALSE, message=FALSE}
# Load data from .RData file
load("/Users/alekhyapidugu/Desktop/Homework 3/submission2/Hmk3_workspace.Rdata")
```

\newpage
# Question 1 <br>  
```{r results="hide", echo=FALSE, message=FALSE}
final.data <- readRDS('/Users/alekhyapidugu/Desktop/Homework 3/data/output/TaxBurden_Data.rds')

final.data <- final.data %>% group_by(state)%>%
arrange(state, Year)%>%
mutate(tax_change = tax_state-lag(tax_state),
tax_change_d= ifelse(tax_change == 0,0,1),
price_cpi_2022 = cost_per_pack*(cpi.2012/index),
total_tax_cpi_2022 = tax_dollar*(cpi.2012/index),
ln_tax_2012  =log(total_tax_cpi_2022),
ln_sales = log(sales_per_capita),
ln_price = log(price_cpi_2022))


q1 <- final.data %>%
  filter(Year >= 1970 & Year <= 1985) %>%  
  group_by(Year) %>%                      
  summarise(count_tax_change = sum(tax_change_d, na.rm = TRUE)) %>%
  mutate(proportion = count_tax_change / n_distinct(final.data$state)) 

# Creating the bar graph
ggplot(q1, aes(x = as.factor(Year), y = proportion)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Proportion of States with Change in Cigarette Tax (1970-1985)",
       x = "Year",
       y = "Proportion of States") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
\newpage
# Question 2 <br>  
```{r results="hide", echo=FALSE, message=FALSE}
average_tax <- final.data %>%
  group_by(Year) %>%
  summarise(avg_tax_2012 = mean(tax_state * (218 / price_cpi)))

# Calculate average price of a pack of cigarettes for each year
average_price <- final.data %>%
  group_by(Year) %>%
  summarise(avg_price = mean(cost_per_pack))

# Combine data into a single dataframe
combined_data <- inner_join(average_tax, average_price, by = "Year")

# Plotting
ggplot(combined_data, aes(x = Year)) +
  geom_line(aes(y = avg_tax_2012, color = "Average Tax (2012 dollars)")) +
  geom_line(aes(y = avg_price, color = "Average Price of a Pack")) +
  scale_color_manual(values = c("Average Tax (2012 dollars)" = "blue", "Average Price of a Pack" = "red")) +
  labs(title = "Average Tax and Price of Cigarettes (1970-2018)",
       x = "Year",
       y = "Average Amount",
       color = "Metric") +
  theme_minimal()
```

\newpage
# Question 3 <br>  
Connecticut  
<br> 
D.C  
<br> 
Massachusetts 
<br>  
New York  
<br> 
Rhode Island  <br>
```{r results="hide", echo=FALSE, message=FALSE}
price_increase <- final.data %>%
  group_by(state) %>%
  summarise(price_increase = max(cost_per_pack) - min(cost_per_pack))

# Identify the 5 states with the highest increases in cigarette prices
top_5_states <- price_increase %>%
  arrange(desc(price_increase)) %>%
  head(5) %>%
  pull(state)

# Filter data for the top 5 states
top_5_data <- final.data %>%
  filter(state %in% top_5_states)

# Plot the average number of packs sold per capita for the top 5 states from 1970 to 2018
ggplot(top_5_data, aes(x = Year, y = sales_per_capita, group = state, color = state)) +
  geom_line() +
  labs(title = "Average Number of Packs Sold Per Capita (1970-2018) for Top 5 States with Highest Price Increases",
       x = "Year",
       y = "Average Number of Packs Sold Per Capita") +
  theme_minimal()
```

\newpage
# Question 4 <br>  
Georgia 
<br> 
Missouri  
<br> 
North Carolina 
<br>  
North Dakota  
<br> 
Tennessee  <br>
```{r results="hide", echo=FALSE, message=FALSE}
bottom_5_states <- price_increase %>%
  arrange(price_increase) %>%
  head(5) %>%
  pull(state)

# Filter data for the bottom 5 states
bottom_5_data <- final.data %>%
  filter(state %in% bottom_5_states)

# Plot the average number of packs sold per capita for the bottom 5 states from 1970 to 2018
ggplot(bottom_5_data, aes(x = Year, y = sales_per_capita, group = state, color = state)) +
  geom_line() +
  labs(title = "Average Number of Packs Sold Per Capita (1970-2018) for Bottom 5 States with Lowest Price Increases",
       x = "Year",
       y = "Average Number of Packs Sold Per Capita") +
  theme_minimal()

```

\newpage
# Question 5 <br> 
One significant trend to observe is that the bottom five states cigarette prices is not as substantial as the magnitude of the increase in cigarette prices for the 5 states with the highest price raises. The bottom 5 states with the lowest price increases generally demonstrated more stable trends in sales per capita. Although there were fluctuations, the overall trend tended to be less pronounced compared to states with higher price increases.

\newpage
# Question 6 <br>  
The estimated price elasticity of demand for cigarettes from 1970 to 1990 is approximately -0.17. The demand for cigarettes during this period is relatively inelastic, suggesting that changes in price have a limited impact on quantity demanded.

\newpage
# Question 7 <br>  
The estimated price elasticity of demand for cigarettes from 1970 to 1990 using an instrument (total cigarette tax) for log prices is approximately -0.28.
When using an instrument, the estimated price elasticity of demand may differ from estimates without an instrument due to the endogeneity of prices. An instrument helps address potential biases in the estimation by providing exogenous variation in prices.

\newpage
# Question 8 <br> 
```{r  echo=FALSE, message=FALSE}
period_data <- final.data %>%
  filter(Year >= 1970 & Year <= 1990)
first_stage <- lm(log(cost_per_pack) ~ log(tax_dollar), data = period_data)

# Display first-stage results
summary(first_stage)

# Obtain predicted values of log prices
period_data$predicted_log_prices <- predict(first_stage)

# Perform second-stage regression: Regress log sales on predicted log prices
second_stage <- lm(log(sales_per_capita) ~ predicted_log_prices, data = period_data)

# Display reduced-form results
summary(second_stage)
```

\newpage
# Question 9 <br> 

The estimated price elasticity of demand for cigarettes from 1991 to 2015 is approximately -0.67. The demand for cigarettes during this period is relatively inelastic, suggesting that changes in price have a limited impact on quantity demanded.
The estimated price elasticity of demand for cigarettes from 1991 to 2015 using an instrument (total cigarette tax) for log prices is approximately -0.76.

```{r  echo=FALSE, message=FALSE}
period_data <- final.data %>%
  filter(Year >= 1991 & Year <= 2015)

# Perform regression of log sales on log prices
demand_regression <- lm(log(sales_per_capita) ~ log(cost_per_pack), data = period_data)

# Extract coefficient for log price
price_elasticity <- coef(demand_regression)[2]


# Perform first-stage regression: Regress log prices on total cigarette tax
first_stage <- lm(log(cost_per_pack) ~ log(tax_dollar), data = period_data)

# Obtain predicted values of log prices
period_data$predicted_log_prices <- predict(first_stage)

# Perform second-stage regression: Regress log sales on predicted log prices
second_stage <- lm(log(sales_per_capita) ~ predicted_log_prices, data = period_data)

# Extract coefficient for predicted log prices
price_elasticity_with_instrument <- coef(second_stage)[2]

# Interpret results
cat("The estimated price elasticity of demand for cigarettes from 1991 to 2015 using an instrument (total cigarette tax) for log prices is approximately", round(price_elasticity_with_instrument, 2), ".\n")


# Perform first-stage regression: Regress log prices on total cigarette tax
first_stage <- lm(log(cost_per_pack) ~ log(tax_dollar), data = period_data)

# Display first-stage results
summary(first_stage)

# Obtain predicted values of log prices
period_data$predicted_log_prices <- predict(first_stage)

# Perform second-stage regression: Regress log sales on predicted log prices
second_stage <- lm(log(sales_per_capita) ~ predicted_log_prices, data = period_data)

# Display reduced-form results
summary(second_stage)
```

\newpage
# Question 10 <br>
The elasticity estimate for the period 1970-1990 was approximately -0.17, whereas for the period 1991-2015, it was approximately -0.67.
The difference in elasticity estimates could be attributed to changes in consumer behavior, regulatory policies, and socio-economic conditions. For example, increased awareness of health risks associated with smoking could have made consumers more sensitive to price changes in later years.